An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-21 EconGrowth.Rnw
Now including US, UK, JP, and CN for CES fits.

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text
\usepackage{amsmath} %Allows \begin{equation*} for unnumbered equations

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  warning=TRUE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<nlmrt tests, eval=FALSE>>=
# From https://stat.ethz.ch/pipermail/r-help/2012-November/330554.html
Laptop_sale <- c(1405,1863,2027,2669,2938,5275,6595,6943,8621,10905,12420,22400,32380,31600,34900,43163,47838,47592)
CuSale <- Laptop_sale
time <- seq_along(Laptop_sale)
Bass.Model <- Laptop_sale ~ m * ((p + q)^2/p) * (exp(-(p + q) * time)/((q /p) * exp(-(p + q) * time) + 1)^2)
# Bass.Fit <- nls(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), trace = FALSE)
# Bass.Fit
library(nlmrt)
xdata<-data.frame(Laptop_sale, time)
Bass.Fitn <- nlxb(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), data=xdata,trace=TRUE)
print("")
print("After nlxb fit:")
Bass.Fitn
print(summary(Bass.Fitn))

Bass.Fitn <- wrapnls(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), data=xdata,trace=FALSE)
print("")
print("After wrapnls fit:")
print(summary(Bass.Fitn))
@

@@

<<ces functions, eval=TRUE>>=
####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
##
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadData(countryAbbrev)
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish default guesses for sigma, gamma_E, beta, lambda_L and lambda_E.
  start <- list(sigma=0.9, gamma_E=0.01, beta=0.7, lambda_L=0.011, lambda_E=0.018) #values for Sweden
  # Establish bounds
  lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-Inf, lambda_E=-Inf)
  upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= Inf, lambda_E= Inf)
  # Now some country-specific values.
  if (countryAbbrev == "US"){
    start <- list(sigma=0.95, gamma_E=0.1, beta=0.7, lambda_L=0.015, lambda_E=0.015) #values for Sweden
  }
  if (countryAbbrev == "UK"){
    start <- list(sigma=0.95, gamma_E=0.2, beta=0.7, lambda_L=0.02, lambda_E=0.02) #values for Sweden
    lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-0.3, lambda_E=-0.3)
    upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= 0.3, lambda_E= 0.3)
  }
  if (countryAbbrev == "CN"){

  }
  # Runs a non-linear least squares fit to the data with constraints
#   model <- iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
#                                                   * iLabor^beta
#                                                   * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
#                   + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
  model <- iGDP ~ ((1-gamma_E) * ((exp(lambda_L*iYear))^beta * iLabor^beta * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
                  + gamma_E * (exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
  modelCES <- nls(formula=model, data=dataTable, start=start,
                  control=nls.control(maxiter=10000, tol=1e-05, minFactor=1/1024, 
                                      printEval=TRUE, warnOnly=TRUE),
                  algorithm = "port",
                  lower = lower,
                  upper = upper,
                  trace=TRUE
                  )
#   modelCES <- nlxb(model,
#                    data=dataTable,
#                    start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
#                               lambda_L=lambda_LGuess, lambda_E=lambda_EGuess),
#                    trace=TRUE
#                   )
  return(modelCES)
}
print("US Q")
print(summary(cesModel("US", "Q")))

print("")
print("UK Q")
print(summary(cesModel("UK", "Q")))

print("")
print("JP Q")
print(summary(cesModel("JP", "Q")))

print("")
print("CN Q")
print(summary(cesModel("CN", "Q")))

print("")
print("ZA Q")
print(summary(cesModel("ZA", "Q")))

print("")
print("SA Q")
print(summary(cesModel("SA", "Q")))

print("")
print("IR Q")
print(summary(cesModel("IR", "Q")))

print("")
print("TZ Q")
print(summary(cesModel("TZ", "Q")))

print("")
print("ZM Q")
print(summary(cesModel("ZM", "Q")))

@@

\citet{Stern:2012ey} developed a CES model of the form

\begin{equation} \label{eq:CES_model}
  y = \left\{ (1-\gamma_E) \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
      + (\gamma_E) \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
      ^{\frac{\sigma}{\sigma - 1}},
\end{equation}

\noindent where $e$ is any of $q$, $x$, or $u$; $\sigma$, $\gamma_E$, $\beta$, $\lambda_L$, and $\lambda_E$ are fitting parameters; $\mathrm{e}$ is the base of the natural logarithm; and the sum $\alpha + \beta$ is constrained to unity.\footnote{The CES equation in \citet{Stern:2012ey} appears as

@@

but the $1/\sigma$ exponent on the $(1-\gamma_E)$ and $\gamma_E$ terms seems erroneous due to $y(t_\mathrm{0}) \neq 1$.}

>>>>

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text
\usepackage{amsmath} %Allows \begin{equation*} \end{equation*} for unnumbered equations

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  warning=TRUE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

@@

<<ces functions, eval=TRUE>>=
####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
##
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadData(countryAbbrev)
  if (energyType == "U"){
    # Trim the dataset to include only those years for which U is available.
    dataTable <- subset(dataTable, !is.na(iU))
  }
  # Trim data from China and South Africa that is missing labor information for 2011.
  dataTable <- subset(dataTable, !is.na(iLabor))
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish default guesses for sigma, gamma_E, beta, lambda_L and lambda_E.
  start <- list(sigma=0.9, gamma_E=0.01, beta=0.7, lambda_L=0.011, lambda_E=0.018) #values for Sweden
  # Establish bounds
  lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-Inf, lambda_E=-Inf)
  upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= Inf, lambda_E= Inf)
  # Now some country-specific values.
  if (countryAbbrev == "US"){
    start <- list(sigma=0.95, gamma_E=0.1, beta=0.7, lambda_L=0.015, lambda_E=0.015) #values for Sweden
  }
  if (countryAbbrev == "UK"){
    start <- list(sigma=0.95, gamma_E=0.2, beta=0.7, lambda_L=0.02, lambda_E=0.02) #values for Sweden
    lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-0.3, lambda_E=-0.3)
    upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= 0.3, lambda_E= 0.3)
  }
  if (countryAbbrev == "CN"){

  }
  # Runs a non-linear least squares fit to the data with constraints
  # This is the version of the function from the Stern and Kander (2012) paper. However, it includes a 
  # 1/sigma exponent on the gamma_E terms, which, I believe, is bogus.
#   model <- iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
#                                                   * iLabor^beta
#                                                   * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
#                   + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))

  # This is the corrected version of the Stern and Kander (2012) paper.
#   model <- iGDP ~ ((1-gamma_E) * ((exp(lambda_L*iYear))^beta * iLabor^beta * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
#                   + gamma_E * (exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))

  # This is the nls fit to the CES model  
#   modelCES <- nls(formula=model, data=dataTable, start=start,
#                   control=nls.control(maxiter=10000, tol=1e-05, minFactor=1/1024, 
#                                       printEval=TRUE, warnOnly=TRUE),
#                   algorithm = "port",
#                   lower = lower,
#                   upper = upper,
#                   trace=TRUE
#                   )

  # An attempt with nlmrt function nlxb. Doesn't provide many good results.
#   modelCES <- nlxb(model,
#                    data=dataTable,
#                    start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
#                               lambda_L=lambda_LGuess, lambda_E=lambda_EGuess),
#                    trace=TRUE
#                   )

  # Try a fit to a nested CES function with the micEconCES package.  
  # This approach is based on a white paper by Henningsen and Henningsen 
  # http://okonomi.foi.dk/workingpapers/WPpdf/WP2011/WP_2011_9_econometric_estimation_CES.pdf.
  # Set a few of the variables up front
  yName = "iGDP"
  xNames = c("iCapStk", "iLabor", "iEToFit")
  tName = "iYear"
  control=nls.lm.control(maxiter=1000, maxfev=2000)
  if (countryAbbrev == "US") {
    # With Q, The unconstrained optimization leads to rho1 = -0.72 and sigma1 = 3.57, which is not 
    # economically meaningful.  So, a grid search was performed, and it confirmed that low values of 
    # rho1 yield a better fit. In fact, the lower the value of rho1, the better. So, we'll set it 
    # here at rho1=0.0, which makes the elasticity of substitution between k and l sigma1 = 1.0, 
    # giving the Cobb-Douglas form for the k l portion of the CES function. Similar results are found
    # for X and U.
    rho1 = 0.0
    # Unconstrained optimization leads to rho = 25.7 and sigma = 0.0374, indicating that (kl) and (e) are 
    # complimentary.  We'll let the estCES function solve for the best value.
    modelCES <- cesEst(data=dataTable, yName=yName, xNames=xNames, tName=tName, control=control,
                       method="LM",
                       rho1=rho1
  } else if (countryAbbrev == "UK") {
    # The unconstrained optimization leads to rho1 = -2.4 and sigma1 = -0.71, which is not economically meaningful.
    # So, a grid search was performed, and it confirmed that low values of rho1 yield a better fit. In fact, the 
    # lower the value of rho1, the better. So, we'll set it here at rho1=0.0, which makes the elasticity of 
    # substitution between k and l sigma1 = 1.0, giving the Cobb-Douglas form for the k l portion of the CES function.
    # rho1 = c(0.00, 0.01, 0.02)
    rho1 = 0.0
    modelCES <- cesEst(data=dataTable, yName=yName, xNames=xNames, tName=tName, control=control,
                       method="LM",
                       rho1=rho1
  } else if (countryAbbrev == "JP") {
    # The unconstrained optimization leads to rho1 = 7.99 and sigma1 = 0.11, indicating complementarity between k & l,
    # but different from the results for the US and UK (where k & l were found to be substitutes).  However, the
    # p value for rho1 is very high (0.3), indicating that rho1 is not making a significant contribution to the model. 
    # The unconstrained fit gives R^ = 0.996 and sigma = 0.103, indicating complementarity between (kl) and (e).
    # In fact, we could force rho1 = 0.0 and achieve rho = 280 and sigma = 0.00356 with R^2 = 0.988, a very small
    # decrease, indeed.  Note that both the constrained and unconstrained fits give small sigma, indicating
    # that (kl) and (e) are complimentary.
    modelCES <- cesEst(data=dataTable, yName=yName, xNames=xNames, tName=tName, control=control,
                       method="LM"
  } else if (countryAbbrev == "CN"){
    # Unconstrained optimization: rho1 < -0.2290, rho = 14.1651, R^2 = 0.9984. This corresponds to sigma1 = 1.297
    # and sigma = 0.0659, indicating that (kl) and (e) are complements and k & l are substitutes.  However, none of the
    # rho or sigma values are statistically significant.
    # Now, sigma1 > 1 is not economically meaningful, so we'll set rho1 = 0 to force sigma1 = 1.0. When we do that, 
    # we obtain sigma1 = 1.0 (so perfect substitution between k and l) and sigma = 7.19 (again, not economically
    # meaningful) with R^2 = 0.9983.  Again, none of the rho or sigma values are statistically significant.
    # Furthermore, the only significant varluable is gamma = 1.0731.
    # So, for the final fit, I'll put rho1 = 0.0 and rho = 0.0. The final R^2 = 
    rho1 = 0.0
    rho = 0.0
    modelCES <- cesEst(data=dataTable, yName=yName, xNames=xNames, tName=tName, control=control,
                       method="LM",
                       rho1=rho1,
                       rho=rho                     

  return(modelCES)
}
# print("US U")
# print(summary(cesModel("US", "U")))
# 
# print("")
# print("UK U")
# print(summary(cesModel("UK", "U")))
# 
# print("")
# print("JP U")
# print(summary(cesModel("JP", "U")))

# print("")
print("CN Q")
print(summary(cesModel("CN", "Q")))
# 
# print("")
# print("ZA Q")
# print(summary(cesModel("ZA", "Q")))
# 
# print("")
# print("SA Q")
# print(summary(cesModel("SA", "Q")))
# 
# print("")
# print("IR Q")
# print(summary(cesModel("IR", "Q")))
# 
# print("")
# print("TZ Q")
# print(summary(cesModel("TZ", "Q")))
# 
# print("")
# print("ZM Q")
# print(summary(cesModel("ZM", "Q")))

@@

\citet{Stern:2012ey} developed a CES model of the form\footnote{The CES equation in \citet{Stern:2012ey} appears as

@@

but the $1/\sigma$ exponent on the $(1-\gamma_E)$ and $\gamma_E$ terms is erroneous due to $y_{\mathrm{t_0}} \neq 1$.}

\begin{equation} \label{eq:Stern_CES_model_in_print}
  y = \left\{ (1-\gamma_E) \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
      + (\gamma_E) \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
      ^{\frac{\sigma}{\sigma - 1}},
\end{equation}

\noindent where $e$ is any of $q$, $x$, or $u$; $\sigma$, $\gamma_E$, $\beta$, $\lambda_L$, and $\lambda_E$ are fitting parameters; $\mathrm{e}$ is the base of the natural logarithm; and the sum $\alpha + \beta$ is constrained to unity.

The R package micEconCES (***** Need reference here *****) estimates CES production functions of the form

\begin{equation} \label{eq:micEconCES_model}
  y = something.
\end{equation}

\citet{Henningsen:2011td}, in their detailed analysis of \citet{Kemfert:1998vd}, found that

\begin{quote}
\ldots the Levenberg-Marquardt and the PORT algorithms are--at least in this study--most likely to find the coefficients that give the best fit to the model, where the PORT algorithm can be used to restrict the estimates to the economically meaningful region.
\end{quote}
